\vskip-0.6in Endogenous Politics and the Design of Trade Institutions

### 2015-01-10 draft010515.tex
complete ex_ante revisions to intro, abstract, conclusion. update to_do_list

\begin{quote}
{\small Political pressure is undoubtedly an important influence in the setting of trade policy and the formulation of trade agreements. Most of the literature models the political pressure that governments face as resulting from an exogenous, stochastic process. This paper shows that when political pressure arises endogenously, important results can be overturned and new insights into the motivation for features of the trade agreements we observe and rules of organizations such as the WTO come to light. Using a weighted Baldwin-style government objective function, I show that governments may want to use tariff caps both both to force special interest groups to continue lobbying after a trade agreement is signed and to reduce the magnitude of that lobbying effort. Endogenous politics can destroy an escape clause's ability to provide flexibility in times of large negative political shocks when lobbies use the flexibility to seek rents. This can explain why use of WTO Safeguards are conditioned on measurable economic indicators as well as why Safeguard levels of protection are not regulated.}
\end{quote}

@@

I introduce a weighted government utility function that is a slight modification of the standard Baldwin-style government objective function to demonstrate that it may be in even a politically-motivated government's interest to use a trade agreement to restrain political activity. The fact that the WTO Safeguards Agreement replaced compensation in the first three years of the invocation of a safeguard with a dynamic use constraint is consistent with the idea that the organization and the governments that constitute it are working to reduce rent seeking by lobbies. Section~\ref{sec:escape} demonstrates that lobbies do not care about the compensation since the costs do not fall on them, whereas a dynamic use constraint directly affects their ability to receive protection in the future.

There are many exciting avenues to extend this work. Incorporating lobbying over the formation of trade agreement would help to build a bridge to the important work at the intersection of endogenous political economy and trade policy. Further, this simple framework demonstrates a tractable way to introduce endogenous political pressure into many of the important questions that have been and are currently being explored concerning the design of trade agreements and the institutions that facilitate them.

In particular, the discussion in Section~\ref{sec:exante} presumes both that there are no significant interactions between lobbies and that governments are free to choose which sectors are part of a trade agreement, but these decisions are part of a more complicated game that is, in particular, influenced by WTO rules. Taking into account the interactions between multiple lobbies, the influence of WTO rules, and differences in the $\ga$ functions across industries---that is, in how political pressure translates into weight in the policy-making process---seem like important and interesting avenues for further study.

>>>>

\begin{quote}
{\small Political pressure is undoubtedly an important influence in the setting of trade policy and the formulation of trade agreements. Most of the literature in this area models the political pressure that governments face as resulting from an exogenous, stochastic process. This paper shows that when political pressure arises endogenously, important results can be overturned and new insights into the motivation for features of the trade agreements we observe and rules of organizations such as the WTO come to light. Using a fully-weighted Baldwin-style government objective function, I show that governments may want to use tariff caps both both to force special interest groups to continue lobbying after a trade agreement is signed and to reduce the magnitude of that lobbying effort. Endogenous politics can destroy an escape clause's ability to provide flexibility in times of large negative political shocks when lobbies use the flexibility to seek rents. This can explain why use of WTO Safeguards are conditioned on measurable economic indicators as well as why Safeguard levels of protection are not regulated.}
\end{quote}

@@

I introduce a fully-weighted government utility function that is a slight modification of the standard Baldwin-style government objective function to demonstrate that it may be in even a politically-motivated government's interest to use a trade agreement to restrain political activity. The fact that the WTO Safeguards Agreement replaced compensation in the first three years of the invocation of a safeguard with a dynamic use constraint is consistent with the idea that the organization and the governments that constitute it are working to reduce rent seeking by lobbies. Section~\ref{sec:escape} demonstrates that lobbies do not care about the compensation since the costs do not fall on them, whereas a dynamic use constraint directly affects their ability to receive protection in the future.

There are many exciting avenues to extend this work. This simple framework demonstrates a tractable way to introduce endogenous political pressure into many of the important questions that have been and are currently being explored concerning the design of trade agreements and the institutions that facilitate them.

The discussion in Section~\ref{sec:exante} presumes both that there are no significant interactions between lobbies and that governments are free to choose which sectors are part of a trade agreement, but these decisions are part of a more complicated game that is, in particular, influenced by WTO rules. Taking into account the interactions between multiple lobbies, the influence of WTO rules, and differences in the $\ga$ functions across industries---that is, in how political pressure translates into weight in the policy-making process---seem like important and interesting avenues for further study.

In particular, here there is no scope for the government to encourage lobbying above the level that is optimal from the special interest group's point of view; adding additional lobbies may create insights to this effect. There is also strong potential to gain insights concerning very realistic dynamics between political and economics shocks and lobbying, especially insofar as they impact questions of enforcement and administered protection.

### 2015-01-10 intro.tex
complete ex_ante revisions to intro, abstract, conclusion. update to_do_list

Various trade agreement and institutional design features have been proposed in the literature that can help to make an escape clause work, since it is not incentive compatible to allow governments to take advantage of an escape clause whenever they wish. If some cost can be associated with the use of the escape clause or a dispute settlement body (DSB) can procure a signal about the strength of $\ga$, it is often advantageous to grant this period of relief from trade commitments if the signal is strong enough.

@@

This is an intuitively appealing story, but the logic can break down in the presence of endogenous political pressure. An escape clause allows a government to apply a higher tariff barrier when it experiences intense political pressure. But if a government gets a free pass from the WTO whenever it feels sufficient political pressure from domestic interest groups, those interest groups have a strong incentive to exert the required level of pressure. I show that, depending on the government's objectives, this extra political pressure may reduce the government's political welfare.

I employ a model that closely follows \Textcite{bs2005}, adding an endogenously-determined element to their exogenously-determined political economy weights. I show that this can explain why the conditions for invoking the WTO Safeguards measure are purely observable economic variables, and why the level of protection governments can choose when invoking a Safeguard is not restricted.

I explore the impacts the elevated levels of political pressure that an escape clause might encourage on governments' political welfare. To do this, I compare welfare under the standard Baldwin-style government welfare function and a similar, weighted version. The slight modification to the government objective function demonstrates that governments may want to use trade agreements to reduce lobbying. Thus, examining this alternative welfare function in combination with endogenous lobbying can provide a bridge between the theoretical literature and the claims of trade policy practitioners that an important role of trade agreements is to rein in protectionist pressure.

This generalizes the result from \Textcite{mrc2007} that trade agreements can be used as a domestic political commitment device. In their setting, the political commitment role of trade agreements is present only when capital has some degree of mobility;  here this restriction is not needed to establish the result. In a further generalization of \Textcite{mrc2007}, I show that one of the uses of tariff caps is to incentivize the lobby to engage in the political process after the trade agreement is in place. The lobby loses all incentive to exert effort under a strong binding and so a government who prefers some level of political engagement from the lobby will not use a strong binding. Again, this does not depend on a particular capital mobility structure, so this, perhaps, provides an explanation for the ubiquity of tariff caps. Combined with the idea that governments can use tariff caps to restrain endogenous political pressure, a story emerges in which governments employ trade agreements to carefully manipulate lobbying incentives in order to maximize their political objectives.

@@

In the next section, I present the model and some preliminary results. Section~\ref{sec:rigid} contains the analysis of rigid tariffs while Section~\ref{sec:escape} explores trade agreements with escape clauses. The political welfare function of the government is examined in depth in Section~\ref{sec:objfcn} before returning to examine the escape clause in an environment with both endogenous and exogenous political pressure in Section~\ref{sec:escape2}. Section~\ref{sec:concl} concludes.

>>>>

Various trade agreement and institutional design features have been proposed in the literature that can help to make an escape clause work, since it is not incentive compatible to allow governments to take advantage of an escape clause whenever they wish. If some cost can be associated with the use of the escape clause or a dispute settlement body (DSB) such as that at the World Trade Organization (WTO) can procure a signal about the strength of $\ga$, it is often advantageous to grant this period of relief from trade commitments if the signal is strong enough.

@@

This is an intuitively appealing story, but the logic can break down in the presence of endogenous political pressure. An escape clause allows a government to apply a higher tariff barrier when it experiences intense political pressure. But if a government gets a free pass from the WTO whenever it feels sufficient political pressure from domestic interest groups, those interest groups have a strong incentive to exert the required level of pressure, eviscerating the escape clause.

This is one example of a design question whose answers are sensitive to assumptions about the endogeneity of political pressure. In order to examine it and others, I build a model that closely follows \Textcite{bs2005}, adding an endogenously-determined element to their exogenously-determined political economy weights. I find that the presence of endogenous political pressure changes the optimal choice of both strong and weak tariff bindings (exact tariff commitments and tariff caps, respectively) in quite significantly ways compared with political pressure that is assumed to be exogenous.

As in \Textcite{mrc2007}, I show that one of the uses of tariff caps is to incentivize the lobby to engage in the political process after the trade agreement is in place. Once the trade agreement is in place, the lobby loses all incentive to exert effort under a strong binding and so a government who prefers some level of ex-post political engagement from the lobby will not use a strong binding. This result extends \Textcite{mrc2007} to the standard case, examined here, where specific factors are completely immobile.

This, perhaps, provides an explanation for the ubiquity of tariff caps. Combined with the idea that governments can use tariff caps to restrain endogenous political pressure, a story emerges in which governments employ trade agreements to carefully manipulate lobbying incentives in order to maximize their political objectives.

In order to see this, I compare welfare under the standard \Textcite{baldwin}-style government objective function and a similar, fully-weighted version of the government's objective function. At issue is the fact that the standard version is everywhere increasing in political pressure. In an environment with endogenous political pressure, this leads inexorably to the conclusion that governments never want to discourage special interest groups from exerting pressure. The slight modification to the government objective function demonstrates that governments may indeed want to use trade agreements to reduce lobbying.\footnote{In the environment with only one lobby in each country, the governments will be unable to encourage lobbying in excess of that which is optimal from the lobby's point of view.}

Thus, examining this alternative welfare function in combination with endogenous lobbying can provide a bridge between the theoretical literature and the claims of trade policy practitioners that an important role of trade agreements is to rein in protectionist pressure.

This example shows that it may be important to consider how the features of the government objective function interact with the question being asked. For some inquiries, such as those that are the focus of \Textcite{ms2011, ms2012a} in which the impact of the political economy parameter on the shape of the objective function is not crucial, there may be no loss in employing the simple form. However, for other questions that directly involve the government maximizing with respect to political pressure, a more general treatment may be prudent.\footnote{Note that, although \Textcite{gh94} provide microfoundations for one particular \Textcite{baldwin}-style objective function, it is a very specific form with fixed weights in which $1+a$ is attached to the surplus of groups who lobby, and $a$ to the surplus of those who don't lobby. This cannot microfound a flexible model as in \Textcite{longvousden} in which weights depend on lobbying activity or in which weights are the result of shocks.}

In a more general setting, I show that, depending on the government's objectives, in some cases extra political pressure would reduce the government's political welfare \textit{and} trade agreements can help them alleviate this pressure. I demonstrate that this is true whether or not lobbying is possible ex ante---that is, whether or not special interests are able to influence the formation of the trade agreement. Even in the face of ex-ante lobbying, a trade agreement that takes the form of tariff caps works to restrict lobbying in many cases, but also to keep lobbyists ``in the game.'' This broadens the result from \Textcite{mrc2007}, in which the political commitment role of trade agreements is present only when capital has some degree of mobility; here it holds with specific factors that are perfectly immobile. I present an extensive-form game that helps us to see that trade agreements serve this role even when ex-ante lobbying is possible. Although ex-ante lobbying removes some degree of autonomy from the government, it also screens out groups for whom the extra stage of lobbying becomes unprofitable.

In another example, I show that the same combination---endogenously-determined political pressure and a general approach to the government's objective function---can explain why the conditions for invoking the WTO Safeguards measure rely on purely observable economic variables, and why the level of protection governments can choose when invoking a Safeguard is not restricted. The model formalizes the intuition that if the WTO's Dispute Settlement Body actually followed the procedures suggested by the literature---that is, it enforced commitments based on signals of the political pressure experienced by governments---pressure groups would simply exert more pressure and the Safeguard clause could not play the escape-providing role that is intended for it.

Through the lens of these examples, one can see that modeling choices concerning both the source of political pressure and the form of governments' political objective function have important impacts on questions of the design of and motives behind trade agreements as well as the institutions that help to enforce them.

@@

In the next section, I present the model and some preliminary results. Section~\ref{sec:rigid} contains the analysis of rigid tariffs while Section~\ref{sec:escape} explores trade agreements with escape clauses. The political welfare function of the government is examined in depth in Section~\ref{sec:objfcn} before returning to examine the escape clause in an environment with both endogenous and exogenous political pressure in Section~\ref{sec:escape2}. Section~\ref{sec:exante} details the extension to ex-ante lobbying. Section~\ref{sec:concl} concludes.

### 2015-01-10 to_do_list.tex
add coates & ludema reference to intro
organize intro modification in to_do_list
complete ex_ante revisions to intro, abstract, conclusion. update to_do_list

\vskip.2in
\begin{itemize}
	\item Conclusion
		\begin{itemize}
			\item perhaps re-emphasize that gov't can't completely control lobbying: can't make it higher than lobby's optimal. But this is b/c there's only one lobby
		\end{itemize}
	\item finish escape clause result
	\item intro
		\begin{itemize}
			\item MRC-style results survive incorporation of ex-ante lobbying. TA still works to restrict lobbying in many cases, but also to keep lobbyists ``in the game''
			\item Baldwin-style gov't welfare function works for a set of questions, but not for this one
				\begin{itemize}
					\item Need to be able to say which predictions/design questions are okay: those that don't change lobbying incentives
				\end{itemize}
			\item In MRC, weight on lobby's profits is $1+a$ no matter what. Doesn't vary in lobby's efforts. GH not really microfoundations for flexible model as in Long and Vousden (1991) / Baldwin (1987) model; microfoundations for more restrictive version with fixed weights ($1+a$ on those who lobby, $a$ on those who don't); then decide $\tau$ to change PS, CS, etc. (p. 481 GH94). Can't get shocks as in this literature
			\item Coates and Ludema (2001)---endogenous (add to list of GH, MRC)
			\item How do I want to sell the paper? In relation to what literature? (what lit does endogenous? what lit doesn't?)
			\item What if the WTO actually \textit{did} what the literature says it does? It wouldn't work
			\item Be more explicit about relationship between $\ga(e)$ and $\ga(s)$
			\item Need to convey the possibilities of this set-up for capturing the real life dynamics of shocks integrated with lobbying dynamics, how that interacts with enforcement and administered protection
		\end{itemize}
\end{itemize}

\vskip.5in
For after this draft
\begin{itemize}
	\item Could escape clause be made to work with some kind of dynamic use constraint?
		\begin{itemize}
			\item When would lobby exert effort to top up?
		\end{itemize}
	\item What units are $\pi(\tau)$ and $e$ measured in? (no numeraire)
	\item existence proofs
\end{itemize}

>>>>

\begin{itemize}
	\item finish escape clause result
		\begin{itemize}
			\item Could escape clause be made to work with some kind of dynamic use constraint?
			\item When would lobby exert effort to top up?
		\end{itemize}
	\item What units are $\pi(\tau)$ and $e$ measured in? (no numeraire)
	\item existence proofs
\end{itemize}
